---
title: "Diagnostic Plots in funbiogeo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic Plots in funbiogeo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

`funbiogeo` contains many plotting functions. They have multiple goals: (1)
propose good default plots to inspect and represent your data, (2) provide easy
plotting methods for complex objects (automatically display maps from spatial
objects), and (3) provide plotting methods for computed functional diversity
metrics in other packages. This vignette explains in detail all
plotting functions available in `funbiogeo`, how to use them and how to
interpret them.

Some of these plotting functions use different data inputs and some of them can deal with species categorization, i.e. 
displaying information by each category of species (family, order, endemism status, etc.). We detail the standard needed inputs in the table below, the "Additional input" column represents needed input that are not the other standard tables specified by the other columns.

```{r child="../man/rmdchunks/_table-plot-functions.Rmd"}
```

Table: A table of available plotting functions in `funbiogeo` ordered
alphabetically with their required arguments. The "Additional input" column
reflects the need for other argument that are not of the standard object type.

Let's first load the package and example datasets included in the `funbiogeo`.

```{r setup}
library("funbiogeo")
library("sf")

data("woodiv_locations")
data("woodiv_site_species")
data("woodiv_traits")
```


## Naming Convention

Like in the rest of the `funbiogeo` package, the functions are named following
certain conventions. For one, to avoid any collision with other packages all the
functions are prefixed with `fb_`. Second all plotting functions begin with
either `fb_plot_*()`, when they are regular plots, or `fb_map_*()` when they
plot maps.

The function names in `funbiogeo` are generally long to be as specific and clear
as possible. So in case of doubt re-read the function name, and what it should
represent should be clear from the name.


## Regular Plots

In this section we will describe what we call "regular plots", i.e.
plots of non-spatial objects (density plots, bivariate plots, lollipop charts, heatmaps, etc.).
We made this distinction because maps have their own specific challenges. All
the default regular plots proposed in `funbiogeo` are quite specific to the
data. In each of the section below we'll summarize what the plots is about, what
are the needed arguments, and how to interpret the output.

The function are all described in their specific subsections in alphabetical
order.


### Distribution of trait coverages across sites

Visualizing the trait coverage of all sites can help isolate which traits may
show consistently low coverage. Also this can help notice if some groups of
sites have higher coverage for certain traits than for others.

You can plot this using the `fb_plot_distribution_site_trait_coverage()`
function. It takes two arguments: `site_species` the site by species
`data.frame` and `species_traits` the species by traits `data.frame`. Note that
internally the function uses
[`ggridges`](https://cran.r-project.org/package=ggridges) to plot the
distributions. This package is thus needed to run this function.

Example:

```{r plot-dist-site-cov}
fb_plot_distribution_site_trait_coverage(woodiv_site_species, woodiv_traits)
```

We see the distribution of species coverage per site (along the x-axis) for each
trait (along the y-axis) and with all traits taken together (shown at the top of
the top on the line `all_traits`). The proportions on the y-axis labels are the
average coverage observed for this trait. We see that we have very similar
average coverage for all sites (average number of species for which we have
not NA trait data), whatever the trait we consider, which is about 100%
of species.

This plot considers the distribution of species across sites instead of focusing
only on the species by traits `data.frame`. This maybe useful to realize, for
example, that the traits of a species occurring very rarely are missing this
wouldn't necessarily translate in low site-level trait coverage.


### Plotting the number of sites by species

The function `fb_plot_number_sites_by_species()` allows to explore the site by 
species `data.frame`. It shows the number (and proportion) of sites occupied by
each species. Its first argument `site_species`, the site by species `data.frame` 
is necessary while the second one `threshold_sites_proportion`, a target 
proportion of sites coverage, is optional.

```{r plot-sites-coverage-species}
fb_plot_number_sites_by_species(woodiv_site_species)
```

The function outputs a dotchart. The number of occupied sites per species is indicated at the 
bottom x-axis, while the top x-axis represents the proportion of occupied sites. 
The left y-axis label species names (in parenthesis) and their rank by increasing 
prevalence. Note that for readability constraints, only a limited number of species are labeled on the y-axis, but they are all displayed as dots on the plot.

Adding the second argument `threshold_sites_proportion` displays a vertical
bar at the target proportion of sites to see how many species occupy more or less than 
the given proportion of sites.

Let's say we want to see the species present in at least 40% of the sites:

```{r plot-sites-coverage-species-thresh}
fb_plot_number_sites_by_species(woodiv_site_species, threshold_sites_proportion = 0.4)
```

The threshold bar helps us get a sense of how many species are present in at least 40% of the sites. It also displays the corresponding number of sites.

### Plotting the number of species per trait

One way to look at the species by traits `data.frame` is to look at
number of species with non-missing trait values for each trait. The
`fb_plot_number_species_by_trait()` function does exactly that. Its first
argument `species_traits`, the species by traits `data.frame`, is necessary
while the second one `threshold_species_proportion`, a target proportion of
species coverage, is optional.

```{r plot-species-coverage-trait}
fb_plot_number_species_by_trait(woodiv_traits)
```

The function outputs a lollipop chart. On the bottom x-axis there is the number
of species covered by the given trait (the top x-axis represents the proportion
of species, which is directly proportional). The y-axis represents each trait.
The dot represents the actual coverage observed with the corresponding proportion
of species written on top. With this figure we see that 70.8% of the species
have a non-missing SLA in the species by trait `data.frame`, while
only 100% have a non-missing plant height.

Adding the second argument `threshold_species_proportion` displays a vertical
bar at the target proportion to easily target traits covering a certain
proportion of species. Let's say we want to see traits available for at least
90% of the species:

```{r plot-species-coverage-trait-thresh}
fb_plot_number_species_by_trait(woodiv_traits, threshold_species_proportion = 0.9)
```

The red dashed vertical line shows the corresponding species coverage with
labels indicating the proportion and corresponding number of species (n = 22).


### Plotting the number of traits per species

You can also display the number of traits available per species. Showing
it for each species would be quite difficult to read, so we decided instead to
represent the number of species having given number of known traits. Also, because
most trait ecologists are interested in multiple traits, we considered nested
proportions of traits: considering the number of species covered by at least one
trait, at least two, etc.

To represent such a plot you can use the function
`fb_plot_number_traits_by_species()`, it uses two arguments. The first
`species_traits` is the species by traits `data.frame`. The second argument is
`threshold_species_proportion` which is optional and corresponds to a certain
threshold proportion of species, so that a line can be added to the plot.

Using it on the included dataset gives:

```{r plot-n-traits-species}
fb_plot_number_traits_by_species(woodiv_traits)
```

The x-axis represents the number of species concerned (bottom x-axis displays
the actual number of species, while top x-axis displays the corresponding
proportion of species). The y-axis shows the number of each traits. Note that
the categories are nested: the set of species with at least 1 trait contains the
set of species with at least 2 traits, and so on and so forth. By definition all
species have 0 or more known traits, but we show that in the plot as a reference to
see how the proportion decreases with increasing number of traits. The
proportion is shown as text above each dot representing it. So for example,
there are 79.2% species with at least 3 non-missing traits.

This plot does not tell us if all species with at least 3 traits share the same
trait combination (there are multiple 3 trait combinations), but it's a first indication.

Let's say we're interested in combinations of traits with at least 50% species
covered. We can use the second argument to show it:

```{r plot-n-traits-species-thresh}
fb_plot_number_traits_by_species(woodiv_traits, threshold_species_proportion = 0.5)
```

Adding this argument displays a vertical bar at the target proportion of species
to easily target number of traits covering a certain proportion of species. The
red dashed vertical line shows the corresponding species coverage with labels
indicating the proportion and corresponding number of species (n = 12)

### Plotting environmental position of sites

To see if our sites are biased environmentally it can be nice to locate them
along environmental variables compared to a full region. For the sake of
simplicity we can focus on two variables against which to compare our sites to a
region. That is what the `fb_plot_site_environment()` function does. It has four
arguments: the first one, `site_locations`, provides the locations of sites as an `sf`
object, `environment_raster` is a
[`terra`](http://cran.r-project.org/package=terra) raster object, the next two
arguments are `first_layer` and `second_layer` which are the names of the two
variables to be extracted from `environment_raster` to make our plot.

From the included dataset we can represent the first 6 sites along total annual
precipitation and mean annual temperature over the full region:

```{r}
# Create the environmental rasters
prec <- system.file("extdata", "annual_tot_prec.tif", package = "funbiogeo")
tavg <- system.file("extdata", "annual_mean_temp.tif", package = "funbiogeo")

layers <- terra::rast(c(tavg, prec))

# Make plot (show environmental position of 6 first sites)
fb_plot_site_environment(head(woodiv_locations), layers)
```

The plot shows the first selected layer as the x-axis and the second one as the
y-axis. The environmental position of the sites are displayed using the big blue
dots, while the light gray pixels are all the environmental variables extracted
from the provided environmental raster.

This figure can show us that the first six sites are actually higher in
temperature and lower in annual precipitation than most other places in our
environmental rasters, with one site has higher precipitation than others.


### Plotting trait coverage for all sites

To select appropriate sites and traits, we can visualize the trait coverage per
site and per trait. This is exactly what is done by
`fb_plot_site_traits_completeness()` which takes `site_species`, the site by
species `data.frame`, and `species_traits`, the species by traits `data.frame`,
as arguments.

We can use it with the included dataset as an example:

```{r plot-site-trait-complete}
fb_plot_site_traits_completeness(woodiv_site_species, woodiv_traits)
```

The plot shows the trait along the x-axis (and their average coverage across all
sites in their labels) and sites along the y-axis. Each thin horizontal line
represents a site. The color indicates the coverage for the trait in each
column. Note that for readability reasons the color scale has been discretized
from 0 to 100% coverage. Traits are ranked in decreasing average coverage. The
last column `all_traits` contains the coverage for all traits taken together.

With this figure we can see that all sites have over 95% coverage for all traits.


### Plotting trait coverage per species

To select appropriate traits, we can visualize the trait coverage per species. 
This is done by `fb_plot_species_traits_completeness()` which takes
`species_traits`, the species by traits `data.frame`, for sole argument.

We can use it with the included dataset as an example:

```{r plot-species-trait-complete}
fb_plot_species_traits_completeness(woodiv_traits)
```

This figure visualizes directly the species by trait `data.frame`. The x-axis
displays the different traits, ranked from left to right in decreasing coverage
order (as indicated in the x-axis labels). The last column `all_traits`
considers all traits taken together. The y-axis represents species in decreasing
coverage order from bottom to top. Each cell thus represents the trait for a
species: blue if the trait is known and red if it is missing.

From this plot we see that a small proportion of species have missing
traits and but taken all together means that we know perfectly only 58.3% of the
species.


### Plotting number of species with missing traits

Maybe you're rather interested in plotting the number of species with known
(or missing) traits in your dataset. This plot is an alternative to the one
from the previous section. It can be done through
`fb_plot_species_traits_missingness()` which takes
`species_traits`, the species by traits `data.frame`, for argument as well as
`all_traits` to know if an additional row should be used to display a summary
for all traits taken together.

```{r plot-species-traits-missingness}
fb_plot_species_traits_missingness(woodiv_traits, all_traits = TRUE)
```

The plot displays the number of species with known and missing traits. It shows
each trait in separate line as a proportional bar chart with the total numbers
included within each bar.



### Displaying traits combinations

Looking at the traits coverage for each species continuously maybe impractical or
difficult. For example, when trying to display thousands of species or when the
trait coverage varies widely across species. One way to reduce the size of the
analyzed (and thus visualized) dataset is instead to count at which frequency
appear the combinations of present/missing traits.

This is done by `fb_plot_trait_combination_frequencies()` which takes two
arguments: `species_traits`, the species by traits `data.frame`, and `order_by`
(equals to either `"number"`, the default option, or `"complete"`), which
influences the order of the plot (more details below).

We can use it with the included dataset as an example:

```{r plot-trait-comb-freq}
fb_plot_trait_combination_frequencies(woodiv_traits)
```

The x-axis represents individual traits ranked in alphabetical order from left
to right. The y-axis represents different combinations. The labels on the y-axis
show both the number and the frequency of each combination. When the cell is
blue it means that the trait is present, when it is red it means the trait is
missing. By default the combinations are ordered by increasing number from top
to bottom, with the most numerous combinations of trait presences at the very
bottom of the graph.

With this graph we can see that we have exactly 14 species (58.3% of the total
number of species provided) with all their traits that are non-missing, while 1 species
has a missing seed mass and a missing wood density.

If we change `order_by` to `"complete"`, the combinations are ordered instead by
the number of trait presents among them:

```{r plot-trait-comb-freq-order}
fb_plot_trait_combination_frequencies(woodiv_traits, order_by = "complete")
```

The most complete combinations are shown at the bottom, while the combinations
with most missing traits are at the top.


### Displaying traits correlations

Most of the plot functions in `funbiogeo` show traits independently of one
another. However, for functional biogeography analyses, trait correlations maybe
very relevant. This is exactly what is done by `fb_plot_trait_correlation()`. It
takes as first needed argument `species_traits`,  the species by traits
`data.frame`. The other arguments are optional and will be passed to
`stats::cor()`.

Note that the function only works with **numerical traits** and with **complete
observations**. It silently removes any species that has any `NA`.

With the included dataset as an example:

```{r plot-trait-corr}
fb_plot_trait_correlation(woodiv_traits)
```

Both x- and y-axes represent the different traits. At their intersection are
shown squared that are colored in function of their correlation coefficient
(purple means close to -1 correlation while brown means close to 1, white means
close to 0). The correlation coefficients are also displayed in the middle of
this square.

With this visualization we can see that our traits are mostly unorrelated, there
is a slight negative correlation of plant and wood density (cor = -0.2).


## Maps

Map functions in `funbiogeo` are here to provide good default visualization
leveraging the spatial information of sites. We know that producing map in R
is challenging. That's why we provide these helper functions.
Of course, these functions are all basics and you either have to customize them
by adding `ggplot2` commands to the returned plots, or to look at their code to
produce similar plots in the way you want.

For example, the functions do not display background maps because it would be
too complex to accommodate for all use cases between quite localized sites up to
global level analyses.

Note that all maps are always in the Coordinate Reference System of the provided
`site_locations` object. You can also tweak that by adding a custom
`ggplot2::coord_sf()` call.


### Mapping traits coverage of sites

The function `fb_map_site_traits_completeness()` maps the trait coverage (proportion of
species, weighted by relative abundance if relevant, with known trait values) of
each site for all traits.

It takes three arguments: `site_locations` (spatial locations of sites),
`site_species` (site by species `data.frame`), and `species_traits` (species by
traits `data.frame`).

If we use the included dataset in `funbiogeo` it gives:

```{r map-trait-complete}
fb_map_site_traits_completeness(woodiv_locations, woodiv_site_species, woodiv_traits)
```

It displays maps of the sites, colored by their trait coverage. Each facet
represents a different trait while the `all_traits` facet consider all traits
together.

From this plot we can see that on average the center of Spain shows lower 
average trait coverage. It may be an indication of higher species richness in
Spain compares to other places which makes achieving high coverages more
difficult. 


### Mapping arbitrary site data

To be the most flexible possible, `funbiogeo` provides a function to map
arbitrary site data whether quantitative or qualitative with the
`fb_map_site_data()` function. The first argument it takes is `site_locations`,
the spatial locations of sites as an `sf` object, the second argument is
`site_data` which is a `data.frame` containing a `"site"` column and data in
additional columns, the third and last required argument is `selected_col` which
should be the name of the column provided in the `site_data` `data.frame` that is going to be used as
a variable to display. All three arguments are required.

For example, if we want to get a map of species richness of the included dataset
we can do the following:

```{r map-site-data-sr}
# Compute Species Richness
site_rich <- fb_count_species_by_site(woodiv_site_species)

# Map of Species Richness
fb_map_site_data(woodiv_locations, site_rich, "n_species")
```

From this map, we can for example see that the French part of our dataset is
the most species-rich compared to the other places.

Now imagine we want to display the category of sites which be "Testing" or
"Training", depending on which set they belong to. We can use the following to
display the map:

```{r map-site-data-cat}
# Generate categories
site_cat <- data.frame(
  site = woodiv_site_species$"site",
  set  = sample(c("Testing", "Training"), nrow(woodiv_site_species), replace = TRUE)
)

# Display them
fb_map_site_data(woodiv_locations, site_cat, "set")
```

### Mapping site rasters

Mapping rasters can be quite cumbersome with R. `funbiogeo` provides a generic
function to map them with `fb_map_raster()`. It only needs as first argument a
`terra` `SpatRaster` object. It can be very useful when visualizing
environmental layers for example. The function will display the raster in the
provided projection.

For example:

```{r}
# Getting the raster
raw <- system.file("extdata", "annual_mean_temp.tif", package = "funbiogeo")
tavg <- terra::rast(raw)

# Mapping the raster
fb_map_raster(tavg)
```

