---
title: "Diagnostic Plots in funbiogeo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostic Plots in funbiogeo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`funbiogeo` contains many plotting functions. They have multiple goals: (1)
propose good default plots to inspect and represent your data, (2) provide easy
plotting methods for complex objects (automatically display maps from spatial
objects), and (3) provide plotting methods for computed functional diversity
metrics in other packages. This vignette is here to explain in detail all
plotting functions available in `funbiogeo`, how to use them and how to
interpret them.

## Naming Convention

Like in the rest of the `funbiogeo` package, the functions are named following
certain conventions. For one, to avoid any collision with other packages all the
functions are prefixed with `fb_`. Second all plotting functions begin with
either `fb_plot_*()`, when they are regular plots, or `fb_map_*()` when they
plot maps.

## Regular Plots

## Maps



```{r setup}
library(funbiogeo)
```
