---
title: "Get started with funbiogeo"
output: rmarkdown::html_vignette
bibliography: refs.bib
vignette: >
  %\VignetteIndexEntry{Get started with funbiogeo}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7,
  out.width = "100%",
  dpi = 100
)
```

The aim of the `funbiogeo` package is to help users streamline the workflows in 
**fun**ctional **biogeo**graphy [@Violle_emergence_2014].
It helps filter sites, species, and traits based on their trait coverages.
It also provides default diagnostic plots and standard tables summarizing input
data. This vignette aims to be an introduction to the most commonly used
functions.

This vignette is a worked through real world example of a functional biogeography 
workflow using the internal dataset provided in `funbiogeo` and derived from the
WOODIV database [@Monnet_2021].

```{r setup}
library("funbiogeo")
library("ggplot2")
library("sf")
```

# Types of data

The functions in `funbiogeo` mostly leverage three different objects:

- the **species x traits** `data.frame`, which describes the traits (in columns)
of the species (in rows) (`species_traits` argument in `funbiogeo` functions);
- the **site x species** `data.frame`, which describes the presence/absence,
the abundance, or the cover of species (in columns) across sites (in rows)
(`site_species` argument in `funbiogeo` functions);
- the **site x locations** object, which describes the physical location of
sites through an `sf` object (`site_locations` argument in `funbiogeo` functions).

*Note: the `site_locations` object must be an `sf` object, as it is now the 
standard package for spatial data in R. To have more information about the sf
package, refer to its
[website](https://r-spatial.github.io/sf/articles/sf1.html). If you want to
learn how to convert your data to an `sf` object, check the [formatting vignette](long-format.html).*

Optionally, an additional dataset can be provided:

- the **species x categories** object, which contains a potential categorization 
of species like taxonomic classes, specific diets, or any arbitrary classification
(`species_categories` argument in `funbiogeo` functions)


# Provided dataset

We are interested in mapping the functional traits of 24 Conifer tree species
of Western European Mediterranean regions.

You can load the example data using the `data(..., package = "funbiogeo")` call:

```{r load-datasets}
data("woodiv_site_species", package = "funbiogeo")
data("woodiv_categories", package = "funbiogeo")
data("woodiv_locations", package = "funbiogeo")
data("woodiv_traits", package = "funbiogeo")
```

In the following sections, we'll describe in detail these provided
objects in the package.


## Species x Traits

This object contains traits values for multiple traits (as columns) for studied
species (as rows). It should be a `data.frame`. The column containing species 
names should be named **"species"** and the other columns should only contain 
trait values.

Note that we'll be talking about **species** throughout this vignette
and in the arguments of `funbiogeo`, but the package doesn't make any assumption
on the biological level. It can be individuals, populations, strains, species,
genera, families, etc. The important fact is that you should have trait data
for the level at which you want to work. This biological level should correspond
between the trait data (`species_traits` argument) and the occurrence/abundance
 (`site_species` argument) you have.

Let's examine the `woodiv_traits` data included in the package:

```{r preview-species-traits, echo = FALSE}
knitr::kable(
  head(woodiv_traits, 4), 
  caption   = "Format of species x traits dataset", 
  align     = c("c", rep("r", 4))
)
```

The first column **`"species"`** contains species names, while the next four
columns contain different traits for all species. 

Let's look at a summary of the trait dataset:

```{r summary-trait-dataset}
summary(woodiv_traits)
```

Note that to use your own species by traits `data.frame`, it should follow
a similar structure with the first column being named **`"species"`** and the
other ones containing traits.


## Site x Species

This object contains species occurrences/abundance/coverage at sites  
of the study area. It is a `data.frame`. The first column, **`"site"`**,
contains site names, while the other columns contain the distribution of each
species across sites.

As mentioned above for the notion of "species" we're referring to, we are
talking about sites in an abstract way. `funbiogeo` doesn't have a specific
notion of sites, but these can be any groupings of individuals whatever the
scale. These could, for example, be plots, assemblages, countries, or biomes.

The package `funbiogeo` comes with the example dataset on Conifer tree species 
of Western European Mediterranean regions `woodiv_site_species`.
Let's look at it:

```{r preview-sites-species, echo = FALSE}
knitr::kable(
  head(woodiv_site_species[ , 1:6], 10), 
  caption   = "Format of site x species dataset", 
  align     = rep("c", 6)
)
```

The example dataset contains the occurrence of the 24 Conifer tree species
across 5,366 sites (grid cells of 10 km x 10 km resolution).

To use your own site by species `data.frame`, you should follow a similar
structure with the first column being named **`"site"`** and the other ones
containing presence information of species across sites.


## Site x Locations

This object contains the geographical location of the sites. It should be an
`sf` object from the [`sf` package](https://cran.r-project.org/package=sf).
These are spatial R objects that describe geographical locations. The sites can
have arbitrary shapes: points, regular polygons, irregular polygons, or even
line transects! To make sure that your data is well plotted you should specify
the Coordinate Reference System (CRS) of this object.

The package `funbiogeo` comes with the example dataset `woodiv_locations`
defining the location of the 5,366 sites (grid cells of 10 km x 10 km
resolution) as polygons. It contains the label of the sites in its first column
**`"site"`**:

```{r preview-sites-locs}
head(woodiv_locations)
```

Note that to use your own site locations object, it should follow a similar
structure, being an `sf` object with the first column being named **`"site"`**.


## Species x Categories

This optional object contains a categorization of species like taxonomic
classes, specific diets, or any arbitrary classification (`species_categories`
argument in some `funbiogeo` functions). It must contain only two columns: the
**`species`** column and the category of species.

The package `funbiogeo` comes with the example dataset `woodiv_categories` 
providing different species categorizations: the species family, the species 
genus, the species endemism, and the cultivated status of the species.

Let's have a look at this dataset:

```{r preview-species-categories, echo = FALSE}
knitr::kable(
  head(woodiv_categories, 5), 
  caption   = "Format of species x categories dataset", 
  align     = c(rep("c", 3), "l", rep("c", 2)),
  row.names = FALSE
)
```

Let's select the **`family`** category:

```{r select-species-categories, echo = FALSE}
woodiv_families <- woodiv_categories[ , c("species", "family")]

head(woodiv_families)
```


# Visualizing the data (diagnostic plots)

`funbiogeo` provides many functions to display the data to help the user select
specific traits, species, and/or sites. We are going to detail some of them in
this section (see the full list and how to interpret them in the
[diagnostic plot vignette](diagnostic-plots.Rmd)). We call them
*diagnostic plots* because they help us to have an overview of our dataset
**prior** to the analyses.

## Trait completeness per species

A first way to visualize our `data.frame` is to look at the proportion of
species with non-missing traits using the `fb_plot_number_species_by_trait()`
function. It takes the species by trait `data.frame` as input.

```{r fig-plot-sp-by-trait}
fb_plot_number_species_by_trait(woodiv_traits)
```

This plot shows us the number of species (along the x-axis) in function of the
trait name (along the y-axis). The number of concerned species is shown at the
bottom of the plot while the corresponding proportion of species (compared to
all the species included in the trait dataset) is indicated as a secondary
x-axis at the top. The proportion of species concerned is shown at the right of
each point. For example, in our example dataset, 70.8% species have a value for
SLA.

The function also includes a way to provide a target proportion of species as
the second argument. It will display the proportion as a dark red dashed line.

For example, if we want to visualize which traits are available for more than
75% of the species (we can also say which traits *cover* more than 75% of the
species), we can use function `fb_plot_number_species_by_trait()`. It takes as
first argument, the species by traits table, then, as second argument the
proportion of species to consider (as a number between 0 and 1):

```{r fig-plot-sp-by-trait-prop}
fb_plot_number_species_by_trait(
  woodiv_traits, threshold_species_proportion = 0.75
)
```

The plot shows us the number (and proportion) of species which have non-missing
values for each trait included in our dataset. The number of species is
specified as the bottom x-axis, while its corresponding proportion is reported
as the top x-axis. According to this graph, we see that plant height is
available for 100% of species, while SLA is only available for 70.8% of the
species in the dataset. The dashed red line represents the proportion we gave as
threshold in the function call above. By the line we see this proportion (75%)
with the corresponding species number (18) as labels on the plot. Given this
plot, we know that only two traits are available for more than 75% of the
species: plant height and seed mass.


## Number of Traits per Species

Another way to filter the data would be to select certain species that
have at least a certain number of traits. This can be visualized using the
`fb_plot_number_traits_by_species()` function. Similarly to the above-mentioned
function, it takes the species x traits `data.frame` as the first argument:

```{r fig-plot-trait-by-sp}
fb_plot_number_traits_by_species(woodiv_traits)
```

The plot shows the number (bottom x-axis) and the proportion (top x-axis) of
species covered by a specific number of traits (0 to 4 in our example). We can
read it as the fact that 58.3% of the species show non-missing values for the
four traits in the dataset. However, all species (100%) have non-missing values
for two or more traits. This doesn't mean that these two traits are the same,
but that all species of datasets have non-missing values for a combination two
traits among the four provided traits.

To identify further which combinations of traits are most frequently available
together, we can use the `fb_plot_trait_combination_frequencies()` function.
This function takes the species-traits table as first argument:

```{r trait-combinations}
fb_plot_trait_combination_frequencies(woodiv_traits)
```

In this plot, each **observed** combination of missing and non-missing trait is
represented as a row. Each column represents a different trait. The y-axis
details how many species (as well as the corresponding proportion) harvest this
particular combination of missing/non-missing trait values. The cells are blue
to represent non-missing trait, and red when they are missing. At the bottom of
the graph, we can indeed see that 14 species (representing 58.3% of the
species), have all traits that are non-missing. At the very top we see that only
one species has the a non-missing plant height and SLA with a missing seed mass
and wood density. This graph allows us to better examine the missingness
patterns among our trait dataset. To see all the available options in
`funbiogeo` as well as the details of the arguments of each function refer to
the [diagnostic plots vignette](diagnostic-plots.html).


# Filtering the data

We performed simple visualizations of our dataset to know identify our patterns
of trait completeness. Based on these plots, we can choose thresholds in trait
completeness to filter our data for our following analyses.

## Filter trait by species coverage

We want to select the traits that are available for at least 75% of the species. 
To do so we can use the `fb_filter_traits_by_species_coverage()` function.
The function takes the species by traits `data.frame` and outputs the same
dataset but with the traits filtered (so with fewer columns). The second
argument `threshold_species_proportion` is the threshold proportion of species
covered:

```{r filter-traits-by-species}
# Initial dimension of the input data
dim(woodiv_traits)

# Filter traits 
selected_traits <- fb_filter_traits_by_species_coverage(
  woodiv_traits, threshold_species_proportion =  0.75
)

dim(selected_traits)

# The reduced data set now has fewer trait columns
head(selected_traits)
```

The function outputs a filtered species-traits dataset retaining only traits
covering at least 75% of the species. In the end, this keep only two traits:
plant height and seed mass.


## Filter species by trait coverage

Now that we obtained a reduced trait dataset, selecting only two traits, this
doesn't mean that these traits are available for all species. We could filter
species that have only non-missing traits through the function
`fb_filter_species_by_trait_coverage()` with the species x traits `data.frame`
as the first argument and the second argument the proportion of traits that
should be known by species. Here, because we want the species to have all the
traits our threshold will be 100%. This way we can filter our dataset again:

```{r filter-species-by-traits}
# Filter species which have at least 100% of known traits
selected_species <- fb_filter_species_by_trait_coverage(
  selected_traits, threshold_traits_proportion = 1
)

dim(selected_species)

head(selected_species)

```

In the end, by filtering the traits available for at least 75% of the species,
have filtering the species that have only non-missing traits for these traits,
we ended with a list of 23 species and 2 traits. This is the dataset we'll
continue using in the rest of the vignette.


## Filter sites by trait coverage

Now that we have filtered our traits and species of interest, we need to filter
the sites that contain enough species for which the traits are available.
Similarly to above the function is `fb_filter_sites_by_trait_coverage()` it
takes as first two arguments the site x species `data.frame` and the species x
traits `data.frame`. The third argument is `threshold_traits_proportion` that
indicates the percent coverage of traits to filter each site. Note that this
coverage is weighted by the occurrence, abundance, or cover depending on the
content of the site x species `data.frame`.

Let's say here we're interested in sites for which our species with available
traits represent at least 90% of the species present:

```{r filter-trait-coverage}
# Initial site x species data
dim(woodiv_site_species)

# Filter sites with at least 90% species covered
filt_sites <- fb_filter_sites_by_trait_coverage(
  woodiv_site_species, selected_species, threshold_traits_proportion = 0.9
)

# Filtered sites
dim(filt_sites)

filt_sites[1:4, 1:4]
```

The output of the function is a site x species `data.frame` with selected sites
and species. Now we selected 5,364 sites out of 5,366, for our 2 traits and 23 
species.


# Computing Functional Diversity metrics

The `funbiogeo` functions helped us filter our data appropriately with enough
available trait information for species and sites. The goal of `funbiogeo` is to
help you analyzing functional trait data and computing functional diversity
indices. These indices capture the diversity of trait values in a set of species,
if you're interested in an introduction to functional diversity indices and how
to analyze them, it's out of scope of this vignette, but you can refer to 
Mammola et al. [-@Mammola_Concepts_2021]. The paper provides a general workflow
to work with trait data.

`funbiogeo` doesn't aim to substitute to all these amazing tools that compute a
diversity of indices with different properties and formulas., however, we can
use the filtered datasets to proceed with our analyses This is where you should
use your preferred packages to compute functional diversity indices like
[`mFD`](https://cran.r-project.org/package=mFD) or
[`funrar`](https://cran.r-project.org/package=funrar). If you're new to the
world of functional diversity analyses, we suggest reading
[`mFD` introductory vignette](https://cmlmagneville.github.io/mFD/articles/mFD_general_workflow.html). It underlines the different steps to compute functional diversity metrics. If you're interested in computing functional rarity indices with `funrar` you canalso refer to [its tutorial](https://rekyt.github.io/funrar/articles/funrar.html).

For the sake of the example, we included a function in `funbiogeo`
to compute Community-Weighted Mean [CWM, @Garnier_Plant_2004] named `fb_cwm()`.
The CWM is the abundance-weighted average trait per site. We'll be using it in
the following section, then we'll then show another example computing functional
diversity indices using the `mFD` package.


## Community-Weighted Mean (CWM)

We're interested to look at the spatial distribution of the average plant height
and seed mass of Conifer tree species. To do so, we can compute the
community-weighted mean of both traits. We'll use the `fb_cwm()` function to do
so, it takes the site x species `data.frame` and species x traits `data.frame`
as arguments.

```{r compute-cwm}
# We're reusing our filtered data, on both the site x species data.frame
# and the species x traits data.frame to compute CWM
cwm <- fb_cwm(filt_sites, selected_species)

head(cwm)
```

It outputs a `data.frame` with 3 columns: the first one, `site`, shows the site
name as provided in the input site x species `data.frame`, `trait` which
indicates the trait name on which the CWM is computed, and `cwm` which shows the
value of the CWM at this site for this trait.


## Compute functional diversity indices

We can also integrate our filtered datasets in other functional diversity computation pipeline. We'll show an example by computing functional richness
with the `mFD` package.

Before computing any functional diversity index, we need to scale and center our
trait values so that they are on comparable scales. For this we can use the
`tr.cont.scale()` function in `mFD`. The function requires that species names
are available as row names instead of having a separate `species` column. We
first give row names and then use the function to scale our traits.

*Note: the following chunk of code is executed only if you have `mFD` installed*

```{r mfd-scaling-traits, eval = require("mFD")}
## To install 'mFD' uncomment the following line
# install.packages("mFD")

# Create a new object
subset_traits <- selected_species

# Add row names
rownames(subset_traits) <- subset_traits[["species"]]

# Remove unused 'species' column
subset_traits <- subset_traits[, -1]

# Scale traits
scaled_traits <- mFD::tr.cont.scale(subset_traits)

# Check that traits are scaled (should have a mean of 0 and SD of 1)
summary(scaled_traits)

# Compare initial trait data to scaled traits
head(selected_species)
head(scaled_traits)
```

We get a data.frame with the two scaled traits and species names as row names.

To compute functional diversity indices with `mFD` we further need a site by
species data.frame, with sites names as row names, and only for the species for
which we have the traits in the species by trait table. We thus transform the
site by species object:

```{r mfd-format-site-species, eval = require("mFD")}
# Copy filtered sites
formatted_site_species <- filt_sites

# Put site names as row names
rownames(formatted_site_species) <- filt_sites[["site"]]
formatted_site_species <- formatted_site_species[, -1]

# Keeping only species for which we have the traits
formatted_site_species <- formatted_site_species[, rownames(scaled_traits)]

formatted_site_species[1:5, 1:5]
```

`mFD` furthermore requires that the site by species object is a `matrix`, we
thus convert it to a matrix:

```{r mfd-site-species-matrix, eval = require("mFD")}
formatted_site_species <- as.matrix(formatted_site_species)
```

We can now compute functional diversity metrics using the `alpha.fd.multidim()`
function from the `mFD` package. We'll be using our two formatted objects
`scaled_traits` as our species by trait table, and `formatted_site_species` as
our site by species table. We'll be computing Functional Dispersion (noted FDis)
as a functional diversity index. It's out of the scope of this vignette to
explain the differences between functional diversity indices, but we recommend
reading Mammola et al. [-@Mammola_Concepts_2021] for a general introduction
about them.

Going back to computing Functional Dispersion with the `alpha.fd.multidim()`
function, we need to use the species-traits table as first argument then the
site-species table as second argument, then the name of the functional diversity
index as a third argument:

```{r mfd-fdis, eval = require("mFD")}
woodiv_fdis <- mFD::alpha.fd.multidim(
  scaled_traits, formatted_site_species, ind_vect = "fdis",
  # remove all of the messages
  details_returned = FALSE, verbose = FALSE
)

head(woodiv_fdis$functional_diversity_indices)
```

We now have a table with several diversity indices computed for each site. This
table contains site names as row names an four columns:

- `sp_richn`, which is the species richness, 
- `fdis`, the Functional Dispersion index,
- `fide_plant_height`, the average plant height of all species of the site,
- `fide_seed_mass`, the average seed mass of all species of the site.

Now, we can use `funbiogeo` again to put these different indices on a map to
better understand our dataset. Mapping functions are the topic of the next
section of this introductory vignette.


# Putting variables on the map

## Mapping diversity indices

As we're interested in putting functional dispersion on a map, we should
slightly transform our site-indices table to include an explicit `site` column
as required by `funbiogeo`:

```{r site-indices, eval = require("mFD")}
# Simplify object
woodiv_fdis <- woodiv_fdis$functional_diversity_indices

# Create 'site' column
woodiv_fdis[["site"]] <- rownames(woodiv_fdis)

# Remove row names
rownames(woodiv_fdis) <- NULL

# Move 'site' column as first column
woodiv_fdis <- woodiv_fdis[, c(5, 1:4)]

head(woodiv_fdis)
```

We get a regular data.frame describing each diversity index with the `site`
column giving the id of the sites. Using `funbiogeo` we can automatically create
a map of these variables by using the `fb_map_site_data()` function. It takes as
first three arguments: (1) the site locations `sf` object, (2) the data.frame
containing the data to be plotted with a column named `site`, (3) the name of
the column to be plotted on the map as a character string (= quoted).
Let's represent the functional dispersion:

```{r map-fdis, eval = require("mFD")}
fb_map_site_data(woodiv_locations, woodiv_fdis, "fdis")
```
We get a map of our sites, colored by Functional Dispersion. The map uses the
default `ggplot2` color scheme. We see that we have higher functional dispersion
in Spain and in the French-Spanish border than in Italy.
Because all of the plotting functions of `funbiogeo` output `ggplot2` objects
they can be customized using usual `ggplot2` syntax:

```{r map-fdis-custom, eval = require("mFD")}
fb_map_site_data(woodiv_locations, woodiv_fdis, "fdis") +
  scale_fill_viridis_c() +
  labs(fill = "Functional Dispersion",
       title = "WOODIV (2 traits: plant height & seed mass)") +
  theme(legend.position = "bottom")
```

We can proceed similarly to display the average plant height per assemblage by specifying the name of the column `fide_plant_height` from our `woodiv_fdis` data.frame:

```{r woodiv-avg-plant-height, eval = require("mFD")}
fb_map_site_data(woodiv_locations, woodiv_fdis, "fide_plant_height") +
  # Additional code to customize our plot
  scale_fill_distiller("Average plant height", palette = "RdYlBu") +
  labs(title = "WOODIV average plant height") +
  theme(legend.position = "bottom")
```

We see that the tallest assemblages are in Italy, while central Spain and North
of Portugal shows the lowest average plant height. You can adapt the above code
on the other available diversity indices.


## Mapping an environmental raster

A common case of analysis in biogeography is to be interested in mapping
environmental variables. If we want to display the environment associated with
our sites of interest, we can leverage environmental raster layers as provided, for example, by
WorldClim [@Fick_WorldClim_2017] or CHELSA [@Karger_Climatologies_2017].
Fortunately, we have access to an example raster of mean annual temperature
through `funbiogeo`.

We ar first going to read the raster using the `terra` package, which is the
reference package to read spatial raster data. If you want to know more about
raster data, we recommend reading the
[dedicated chapter](https://r.geocompx.org/raster) in the
[*Geocomputations with R*](https://r.geocompx.org/) book
[@Lovelace_Geocomputation_2025]. Then, we'll use the `fb_map_raster()`
function that displays a map for the first layer of the raster data.
It takes the actual raster object as first argument. The other arguments are
passed to the `theme()` function of ggplot2 to customize the plot.

So first, let's read the mean annual temperature raster provided by `funbiogeo`.
For this, we'll use the `system.file()` function which allows accessing specific
files from packages, then we'll read the raster with the `rast()` function from
the `terra` package:

```{r map-tavg}
# Get file pathr
tavg <- system.file(
  "extdata", "annual_mean_temp.tif", package = "funbiogeo"
)

# Read raster
tavg <- terra::rast(tavg)

# Display object
tavg
```

This raster represents mean annual temperature in Europe. We can see that the
temperature goes between -5.9°C and 19.8°C. The raster isn't projected (as given
by its [EPSG](https://en.wikipedia.org/wiki/EPSG_Geodetic_Parameter_Dataset)
code).

We then display it with the `fb_map_raster()` function from `funbiogeo` in the
next chunk. 

```{r map-tavg, dev = 'png'}
# Map raster
fb_map_raster(tavg)
```

As with all plots provided by `funbiogeo`, the `fb_map_raster()` returns a
ggplot2 object which can be customized providing additional functions:

```{r map-tavg-custom, dev = 'png'}
# Map raster
fb_map_raster(tavg) + 
  scale_fill_distiller("Temperature", palette = "Spectral") +
  labs(title = "Mean annual temperature in Europe") +
  theme(legend.position = "bottom")
```

To learn about the ggplot2 syntax and functions check the
[`ggplot2` introductory vignette](https://ggplot2.tidyverse.org/articles/ggplot2.html).

Leveraging the `patchwork` package, that allows to combine different `ggplot2`
objects, we can have side-by-side, the map of mean annual temperature and the
one on annual total precipitation:

```{r composition-map, dev = 'png'}
library("patchwork")

# Read raster ------------------------------------------------------------------
tavg <- system.file("extdata", "annual_mean_temp.tif", package = "funbiogeo")
tavg <- terra::rast(tavg)

prec <- system.file("extdata", "annual_tot_prec.tif", package = "funbiogeo")
prec <- terra::rast(prec)

# Individual Maps --------------------------------------------------------------
map_temperature <- fb_map_raster(tavg, legend.position = "none") + 
  scale_fill_distiller("Temperature", palette = "Spectral")

map_precipitation <- fb_map_raster(prec) + 
  scale_fill_distiller("Precipitation", direction = 1)

# Plot composition -------------------------------------------------------------

(map_temperature / map_precipitation) + 
  plot_annotation(title = "Europe", 
                  theme = theme(plot.title = element_text(face = "bold"))) & 
  theme_classic() &
  theme(text = element_text(family = "mono"))
```

This function allows the visualization of a raster in a simple fashion, but it
doesn't tell us anything about the environmental variable at the sites we're
interested in. In the next section we will map an environmental variable at each
site of our study.


## Map of average environmental variable in site

To get the average environmental variable in the site we can use the
`fb_get_environment()` function, it takes as arguments the site-locations object
and an environmental raster from the `terra` package.
By default, it takes the average of the raster values per site.

```{r get-environment}
site_mat <- fb_get_environment(woodiv_locations, tavg)

head(site_mat)
```

The variable names in the columns are based on the names of the provided raster.
To put these values on the map we can use the `fb_map_site_data()` function,
which allows mapping arbitrary site-level variables. It takes three needed
arguments: the first of which, `site_locations`, which is the `sf` object
describing sites' geographic locations; the second argument is `site_data`,
which is a `data.frame` giving additional data indexed by site; and
`selected_col`, which is a character giving the name of the column to plot from
the `site_data` argument.

We can thus plot the mean annual temperature of sites through the following
commands:

```{r map-avg-environment}
fb_map_site_data(woodiv_locations, site_mat, "annual_mean_temp") +
  labs(title = "Mean Annual Temperature per Site")
```



# Conclusion

This concludes our tutorial to introduce `funbiogeo`. The package contains many
more features, especially diagnostic plots, which are explained in detail in
[a dedicated vignette](diagnostic-plots.Rmd).
There is also a specific vignette about
[transforming raw data from long to wide format](long-format.Rmd).
Finally, if you're interested in learning about up-scaling your sites, which
means aggregating your sites at coarser scales,you can
refer to [the specific vignette](upscaling.Rmd)


# References
