---
title: "Formatting your data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formatting your data from raw data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  out.width = "100%"
)
```



```{r setup}
library(funbiogeo)
```


The `funbiogeo`package requires that information is structured in three different datasets:

- the **species x traits** `data.frame` (`species_traits` in `funbiogeo`)
- the **site x species** `data.frame` (`site_species` in `funbiogeo`)
- the **site x locations** object (`site_locations` in `funbiogeo`)



## Wide vs long format



In `funbiogeo` these three datasets **must be** in a wide format (=where one row hosts several variables across columns), but sometimes information is structured in a long format 
(=one observation per row, also called [**tidy** format](https://r4ds.had.co.nz/tidy-data.html)).


For instance, the following dataset illustrates the wider format 
(the presence/absence of all species is spread across columns).


```{r wide-format, echo=FALSE}
wide_data <- data.frame("site"      = LETTERS[1:3],
                        "species_1" = c(1, 0, 1),
                        "species_2" = c(0, 0, 1),
                        "species_3" = c(1, 1, 1),
                        "species_4" = c(1, 1, 0))

knitr::kable(wide_data, caption = "Wide format dataset (used in `funbiogeo`)",
             align = rep("c", ncol(wide_data)))
```


The following dataset illustrates the long format (the column `species` contains the
name of the species and the column `occurrence` contains the presence/absence of species).


```{r long-format, echo=FALSE}
long_data <- data.frame("site"       = rep(LETTERS[1:3], 4),
                        "species"    = c(rep("species_1", 3), rep("species_2", 3), 
                                         rep("species_3", 3), rep("species_4", 3)),
                        "occurrence" = c(1, 0, 1, 0, 0, 1, 1, 1, 1, 1, 1, 0))

knitr::kable(long_data, caption = "Long format dataset",
             align = rep("c", ncol(long_data)))
```



## The `fb_format_*()` functions



If your data are not split into these three wider datasets, you can use the 
functions `fb_format_*()` to create these specific objects from a long format 
dataset.

- The function `fb_format_site_locations()` allows to extract the 
**site x locations** information from the long format data
- `fb_format_site_species()` allows to extract the 
**site x species** information from the long format data
- `fb_format_species_traits()` allows to extract the 
**species x traits** information from the long format data

All these functions take a long dataset as input (argument `data`), where one 
row corresponds to the occurrence/abundance/coverage of one species at one site 
and output a wider object.



## Usage

`funbiogeo` provides a small excerpt of long format data to show how to use the functions.
This data sits at `system.file("extdata", "raw_mammals_data.csv", package = "funbiogeo")`

Let's import the long format dataset provided by `funbiogeo`:

```{r 'load-raw-dataset'}
## Define the path to long format dataset ----

file_name <- system.file("extdata", "raw_mammals_data.csv", package = "funbiogeo")


## Read the file ----

all_data <- read.csv(file_name)
```


```{r preview-raw-dataset, echo=FALSE}
knitr::kable(head(all_data, 10), 
             caption   = "Long table example", 
             align     = c("c", "r", "r", "l", "c", "r", "r", "r", "r", "r", "r")
)
```


### Extracting species x traits data

The function `fb_format_species_traits()` extracts species traits values from 
this long table to create the species x traits dataset. Note that one species 
must have one unique trait value (no trait variation across sites is allowed).

```{r 'format-species-traits'}
## Extract species x traits data ----

species_traits <- fb_format_species_traits(
  data    = all_data, 
  species = "species", 
  traits  = c("adult_body_mass", "gestation_length", "litter_size",
              "max_longevity", "sexual_maturity_age", "diet_breadth")
)

## Preview ----

head(species_traits, 10)
```



### Extracting site x species data

The function `fb_format_site_species()` extracts species 
occurrence/abundance/coverage from this long table to create the 
site x species dataset. Note that one species must have been observed one time 
at one site (the package `funbiogeo` does not yet consider temporal survey).

```{r 'format-sites-species'}
## Format site x species data ----

site_species <- fb_format_site_species(data       = all_data, 
                                       site       = "site", 
                                       species    = "species", 
                                       value      = "count",
                                       na_to_zero = TRUE
)

## Preview ----

head(site_species[ , 1:8], 10)
```



### Extracting site x locations data

The function `fb_format_site_locations()` extracts sites coordinates from this
long table to create the site x locations dataset. Note that one site must have
one unique longitude x latitude value.


```{r 'format-sites-locs'}
## Format site x locations data ----

site_locations <- fb_format_site_locations(data       =  all_data, 
                                           site       = "site", 
                                           longitude  = "longitude", 
                                           latitude   = "latitude",
                                           na_rm      = FALSE)

## Preview ----

head(site_locations)
```


Once your data are in the good format, you can 
[get started](https://frbcesab.github.io/funbiogeo/articles/funbiogeo.html) 
with `funbiogeo`.
