---
title: "Formatting your data from raw data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formatting your data from raw data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(funbiogeo)
```

If your data are not split into three tables you can use the functions
`fb_format_*()` to create these specific objects. These functions take a long
table as input, where one row corresponds to the occurrence/abundance/coverage 
f one species at one site and output a tidy object. Similarly, if your data is
in a long format some variables are repeated across the table
(e.g. sites coordinates, species traits).

For example, the raw data used to generate the internal data of `funbiogeo`:

```{r 'load-raw-dataset'}
## Path to example raw dataset ----
filename <- system.file(
  "extdata", "raw_mammals_data.csv", package = "funbiogeo"
)

## Read CSV file ----
all_data <- read.csv(filename)
```

```{r 'preview-raw-dataset', echo = FALSE}
knitr::kable(
  head(all_data, 10), 
  caption   = "Long table example", 
  align     = c("c", "r", "r", "l", "c", "r", "r", "r", "r", "r", "r")
)
```


## Formatting species x traits data

The function `fb_format_species_traits()` extracts species traits values from 
this long table to create the species x traits dataset. Note that one species 
must have one unique trait value (no trait variation across sites allowed).

```{r 'format-species-traits'}
## Format species x traits data ----
species_traits <- fb_format_species_traits(
  data    = all_data, 
  species = "species", 
  traits  = c("adult_body_mass", "gestation_length", "litter_size",
              "max_longevity", "sexual_maturity_age", "diet_breadth")
)

## Preview ----
species_traits
```

## Formatting site x species data

The function `fb_format_site_species()` extracts species 
occurrence/abundance/coverage from this long table to create the 
site x species dataset. Note that one species must have been observed one time 
at one site (the package `funbiogeo` does not yet consider temporal survey).

```{r 'format-sites-species'}
## Format site x species data ----
site_species <- fb_format_site_species(data       = all_data, 
                                       site       = "site", 
                                       species    = "species", 
                                       value      = "count",
                                       na_to_zero = TRUE)

## Preview ----
dim(site_species)
head(site_species)[,1:10]
```


## Formatting site x locations data

The function `fb_format_site_locations()` extracts sites coordinates from this
long table to create the site x locations dataset. Note that one site must have
one unique longitude x latitude value.

```{r 'format-sites-locs'}
## Format site x locations data ----
site_locations <- fb_format_site_locations(data       =  all_data, 
                                           site       = "site", 
                                           longitude  = "longitude", 
                                           latitude   = "latitude",
                                           na_rm      = FALSE)

## Preview ----
head(site_locations)
```
