---
title: "Upscaling your data: aggregating at coarser scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Upscaling your data: aggregating at coarser scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```


`funbiogeo` provides an easy way to upscale your site data to a coarser
resolution. The idea is that you have any type of data at the site level
(diversity metrics,  environmental data, as well as site-species data) that you
would like to work on or visualize at a coarser scale. The aggregation process
can look daunting at first and be quite difficult to run.
We explain in details, throughout this vignette, how to do so with the
`fb_aggregate_site_data()` function. We'll detail three use cases:

1. Aggregating through a regular square grid
2. Aggregating to irregular polygons (like countries, biomes, or any other shape
)
3. Aggregating using a raster grid

In all three cases, the aggregation can work on any data available at the
site-level. This could be species richness, site-species data, or functional
diversity metrics.


# Preamble: getting site-level data

```{r setup}
library("funbiogeo")
library("sf")
library("ggplot2")
```

Let's import our site by locations object, which describes the geographical
locations of sites.


```{r import-site-locations}
data("woodiv_locations")

woodiv_locations
```

These sites are a collection of regular spatial polygons at a resolution of
10 km x 10 km over South Western Europe. Our site by locations object is an `sf`
object, which means it's a spatial object coupled with a data.frame.

For each site, we want to compute the species richness. We can do so by counting
the species at each site with the function `fb_count_species_by_site()`.

```{r compute-richness}
# Import site x species data
data("woodiv_site_species")

# Compute species richness
species_richness <- fb_count_species_by_site(woodiv_site_species)

head(species_richness)
```

Unfortunately, the `fb_count_species_by_site()` doesn't output a spatial object
but a data.frame with three columns: the identifier of the site, the number of
species, and the proportion of species present at a given site. Before going any
further let's put species richness on the map with the function
`fb_map_site_data()`. That function allows us to represent any arbitrary data
at the site level by combining it with the site by locations object, and it
shows a map of that data:


```{r map-original-richness, dev='png'}
fb_map_site_data(woodiv_locations, species_richness, "n_species")
```

Now, from this site-level species richness, we would like to get the species
richness at different scales. First, we'll show how to aggregate on a coarser
square grid, then at the country level, and then with a grid from a
`SpatRaster`.


# Aggregating through a square grid

Our initial sites are 10km by 10km, we want to aggregate using a grid with
pixels of 300km by 300km. When aggregating with a such a grid, there are two
possibilities: either you already have a defined grid as a spatial object or you
generate one on the fly. Here we'll generate the grid on the fly using the
function `st_make_grid()` from the `sf` package.

Because our data are using a [map projection](https://en.wikipedia.org/wiki/Map_projection) with units in meters,
we can directly specify the `cellsize` argument of `st_make_grid()`.
The function the locations object as first argument, then a vector of two
nubmers as the second argument defining the cell sizes. By default, the newly
created grid will cover the same extent as in the input spatial data.

```{r make-grid}
coarser_grid = sf::st_make_grid(woodiv_locations, cellsize = c(300e3, 300e3))

head(coarser_grid)
```

Because the coarser grid is an `sfc` object, we need to transform it to an `sf`
object to be fully compatible with `fb_aggregate_site_data()`. We do so in the
next chunk:

```{r convert-grid-to-sf}
coarser_grid = sf::st_as_sf(coarser_grid)

coarser_grid
```

Now that we have a new grid, we can plot it with the original site locations
data to make sure it is sensible, compared to what we wanted. We will do so
using the `geom_sf()` function from `ggplot2`.

```{r map-grid}
ggplot() +
  geom_sf(data = coarser_grid, aes(fill = "coarser")) +
  geom_sf(data = woodiv_locations, aes(fill = "original")) +
  # Color the grids
  scale_fill_manual(
    values = c(original = "darkorange", coarser = "darkmagenta")
  ) +
  theme_bw()
```

In orange, you see the original data grid of the data, while in purple is the
coarser grid we defined over which we'll aggregate the data.

We're interested in knowing the occurrence of our Mediterranean plant species
over each pixel of the coarser grid. To do this we'll use the
`fb_aggregate_site_data()` function from `funbiogeo`. It takes as first argument
the site-locations object (including the column with the site ids), then the
data that needs to be aggregated (with the site ids column), then the grid over
which to aggregate the data, finally the last argument is the function to use
to aggregate the data. By default, uses the `mean()` function and assumes
quantitative site data only.

Because we're interested in species richness, we'll use the dataset we
computed in the first section, that defines species richness at each site. We'll
use the `fb_aggregate_site_data()` function on our original site-locations
object, with the species richness data, on the coarser grid, using the `mean()`
function to get back the average species richness across sites per larger pixel.

```{r square-grid-aggregate}
coarser_richness = fb_aggregate_site_data(
  woodiv_locations, species_richness, coarser_grid, fun = mean
)

head(coarser_richness)
```

Note that the object we obtain is also an `sf` object of the same type as the
`grid` we provided.

We can plot it using `ggplot2`:

```{r square-map-richness}
ggplot(coarser_richness, aes(fill = n_species)) +
  geom_sf() +
  scale_fill_viridis_c() +
  theme_bw()
```

We see that the pixels of the North show the greatest average species richness,
especially in the South of France.



# Aggregating through irregular polygons

We aggregated our richness data on a regular grid with square pixels, but it is
very common in biogeography to want to aggregate data on spatial objects that
are of irregular shapes. For example, aggregating data at the biome or the
country scale. Initially, our data may be on a regular grid, but we want to
aggregate at the country scale. This is easily doable with the
`fb_aggregate_site_data()` function.

The function actually works with any arbitrary `sf` object as aggregation grid
whether it's polygons regular or not, lines, points, or a mix of any geometries.

To show you how to do it, we'll use the included spatial object for the four
Mediterrannean countries covered by our example dataset:

```{r coutries-load}
# Load the file from the package folder
countries <- readRDS(
  system.file("extdata", "countries_sf.rds", package = "funbiogeo")
)

# Visualize the spatial object
ggplot(countries) +
  geom_sf(aes(fill = name)) +
  geom_sf(data = woodiv_locations, alpha = 1/2) +  # Add original data
  geom_sf_text(aes(label = name)) +
  theme_bw()
```

As we already have our aggregation object, here the countries delimitation,
we can directly use the `fb_aggregate_site_data()` function on: the original
locations, the species richness data, and the countries object, with the
`mean()` function. As such, we'll obtain the average species richness in
Mediterrannean tree species per country. Note that all the sites that are not
covered by the aggregation polygons (here the countries) are going to be
dropped.

```{r countries-agg}
# Compute country-level average species richness
countries_richness <- fb_aggregate_site_data(
  woodiv_locations, species_richness, countries, fun = mean
)

# Map country-level average species richness
ggplot(countries_richness) +
  geom_sf(aes(fill = n_species)) +
  theme_bw()
```
We observe that France has a higher species richness than the other countries,
with more than 5 species on average in the sites it harbors.


# Aggregating through a SpatRaster grid

In the two previous sections we saw how to aggregate on arbitrary shaped
polygons. However, another quite common use case, especially when working with
species distribution model, is to want to aggregate data follow an environmental
raster or a raster grid.

We would now like to aggregate our sites at the same resolution as is available
for our environmental data such as mean annual temperature.
We need to aggregate site data on a `SpatRaster` spatial grid (from
the [`terra`](http://cran.r-project.org/package=terra) package).
A nice property of `fb_aggregate_site_data()` function is that it outputs
a matching type of object as the provided grid. If it's an `sf` object than the
function will output the same type of `sf` object, while if it's a `SpatRaster`
then it will give back a `SpatRaster` object.

First of all, we can create a coarser grid based on our locations object.


```{r import-grid}
# Import study area grid
coarser_grid <- system.file("extdata", "grid_area.tif", package = "funbiogeo")
coarser_grid <- terra::rast(coarser_grid)

coarser_grid
```

We will aggregate the site data (resolution of 10 km by 10 km) to this new
coarser raster (resolution of 0.83°, close to 92 km by 92km)  with the function `fb_aggregate_site_data()`. This
function requires the following arguments:

- `site_locations`: the site x locations object
- `site_data`: a `matrix` or `data.frame` containing values per sites to
aggregate on the provided grid `agg_grid`. Can have one or several columns
(variables to aggregate). The first column must contain sites names as provided
in the object `species_richness`
- `agg_grid`: a `SpatRaster` object (package `terra`). A raster of one single
layer, that defines the grid along which to aggregate
- `fun`: the function used to aggregate sites values when there are multiple
sites in one cell (do we want to get the minimum value? the maximum? the sum?
or the mean?)

Let's compute our average species richness values across our grid.


```{r upscaled-richness}
# Upscale to grid ----
upscaled_richness <- fb_aggregate_site_data(
  site_locations = woodiv_locations[ , 1, drop = FALSE],
  site_data      = species_richness[ , 1:2],
  agg_grid       = coarser_grid,
  fun            = mean
)

upscaled_richness
```

We get a `SpatRaster` object that is of the same resolution of our provided
`agg_grid` raster grid. The cells of this raster contain the averaged values of
species richness of our sites aggregated on the coarser grid. We can plot these
values through a call to `fb_map_raster()` which allows to plot rasters.

```{r map-upscaled-richness, dev='png'}
fb_map_raster(upscaled_richness)
```


# Specific examples

## Coarsening site-species data

Through the `fb_aggregate_site_data()` function we can also coarsen our
site-species grid by selecting the appropriate function as the `fun` argument,
we detail how in this section.

Now that we've learned how to aggregate arbitrary data at the site scale over a
spatial scale. We're going to use our provided example named `site_species` at a 
resolution of 10 x 10 km to get new sites from a grid with pixels of 0.83° of resolution.

As shown in the previous section, we'll need three objects: `site_species`, which
describes the species present across sites; `site_locations`, which gives the
spatial locations of sites; and `agg_grid` which is a `SpatRaster` object
defining the coarser grid.

We'll use the previously defined object to run our example. To aggregate 
the presence-absence of species within each pixel of the new grid, we'll use
the `max()` function (as the `fun` argument). As such, coarser pixels which
contains a mix of presence and absence of certain species, we'll be considered
as having the species present. Only when the species is absent from all of the
finer scale sites will the coarser pixel show the species as absent.

```{r upscale-site-species}
site_species_agg <- fb_aggregate_site_data(
  woodiv_locations[ , 1, drop = FALSE],
  woodiv_site_species,
  agg_grid = coarser_grid,
  fun = max
)
```

The return object is a `SpatRaster` as well but can be transformed easily in a
data.frame to follow back with the regular analyses provided in `funbiogeo`.
The new object contains one layer for each aggregated variable, i.e. here, one
per species.

```{r show-upscale-site-species}
site_species_agg
```

We can visualize both maps for a single species to see the difference in
resolution :

```{r plot-map-upscale-site-species}

single_species <- merge(
  woodiv_locations, woodiv_site_species[, 1:2], by = "site", all = TRUE
)

finer_map <- ggplot(single_species) +
  geom_sf(aes(fill = as.factor(AALB))) +
  labs(fill = "Presence of AALB", title = "Original resolution (10 x 10 km)")

coarser_map <- fb_map_raster(site_species_agg[[1]]) +
  scale_fill_binned(breaks = c(0, 0.5, 1)) +
  labs(title = "Coarser resolution (0.83°)")

patchwork::wrap_plots(finer_map, coarser_map, nrow = 1)
```


### Obtaining back a site x species `data.frame`

Now we obtained a raster of aggregated site-species presences. However, the 
other functions of `funbiogeo` don't play well with raster data. They need
data.frames to work well. We can do this through the specific function
`as.data.frame()` in `terra` (make sure to check the dedicated help page that
specifies all the additional arguments with `?terra::as.data.frame`).

```{r upscaled-site-species-df}
# Use the 'cells = TRUE' argument to index results with a new cell column
# corresponding to the ID of the coarser grid pixels
site_species_agg_df <- terra::as.data.frame(site_species_agg, cells = TRUE)

site_species_agg_df[1:4, 1:4]

colnames(site_species_agg_df)[1] <- "site"
```

With this, we're ready to reuse all of `funbiogeo` functions to work on these
coarser data. You can proceed similarly to aggregate the ancillary site-related
data, to use them in the rest of the analyses.


## Upscaling functional diversity data

Because `funbiogeo` focuses on the functional biogeography workflow, we'll
explore in this section how to aggregate the results for a functional
biogeography function. First, we'll detail an example aggregating the
community-weighted mean (CWM) of plant height, that is the abundance-weighted
trait average of the assemblage. Second, we'll show an example of coarsing 
functional diversity metrics computed through the [`fundiversity` package](https://funecology.github.io/fundiversity).


### Coarsen CWM of plant height

To compute the CWM we'll use the function `fb_cwm()`.

```{r cwm-body-mass}
site_cwm <- fb_cwm(woodiv_site_species, woodiv_traits[, 1:2])

head(site_cwm)
```

Now we can aggregate the CWM of plant hieght at coarser scale using
`fb_aggregate_site_data()` as done in the previous sections, this time using
the default `fun` argument as we want to compute the average CWM:

```{r upscaled-cwm}
colnames(site_cwm)[3] <- "plant_height"

upscaled_cwm <- fb_aggregate_site_data(
  woodiv_locations[ , 1, drop = FALSE],
  site_cwm[, c(1, 3)],
  coarser_grid
)

upscaled_cwm
```

We can then map the CWM using the `fb_map_raster()` function:

```{r map-upscaled-cwm}
fb_map_raster(upscaled_cwm) +
  scale_fill_continuous(trans = "log10")
```


### Coarser FRic through `fundiversity`

In a similar fashion as in the
[introduction vignette to `funbiogeo`](funbiogeo.Rmd) in this section we'll
compute the Functional Richness using two traits across our example dataset.

```{r compute-fric, eval = require("fundiversity")}
# Get all species for which we have both adult body mass and litter size
subset_traits <- woodiv_traits[
  , c("species", "plant_height", "seed_mass")
]
subset_traits <- subset(
  subset_traits, !is.na(plant_height) & !is.na(seed_mass)
)

# Transform trait data
subset_traits[["plant_height"]] <- as.numeric(
  scale(log10(subset_traits[["plant_height"]]))
)

subset_traits[["seed_mass"]] <- as.numeric(
  scale(subset_traits[["seed_mass"]])
)

# Filter site for which we have trait information for than 80% of species
subset_site <- fb_filter_sites_by_trait_coverage(
  woodiv_site_species, subset_traits, 0.8
)

subset_site <- subset_site[, c("site", subset_traits$species)]

# Remove first column and convert in rownames
rownames(subset_traits) <- subset_traits[["species"]]
subset_traits <- subset_traits[, -1]

rownames(subset_site) <- subset_site[["site"]]
subset_site <- subset_site[, -1]

# Compute FRic
site_fric <- fundiversity::fd_fric(
  subset_traits, subset_site
)

head(site_fric)
```

We can now follow a similar upscaling process as in the previous sections to
compute the average functional richness at a coarser spatial scale:

```{r upscale-fric, eval = require("fundiversity")}
agg_fric <- fb_aggregate_site_data(
  woodiv_locations[ , 1, drop = FALSE], 
  site_fric, 
  coarser_grid
)

fb_map_raster(agg_fric)
```
