---
title: "{{title}}"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::html_document2: 
    toc: true
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_height: 7
    fig_caption: true
    keep_md: false
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: false
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 7
    fig_caption: true
    keep_md: false
  bookdown::word_document2:
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 7
    fig_caption: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width  = 7,
  fig.height = 7,
  dpi        = 100,
  out.width  = "100%",
  echo       = FALSE,
  eval       = TRUE
)

library("funbiogeo")
```

This is an automated report on your species-traits, site-species, and
site-locations data generated with `funbiogeo`. Please find below default analyses and visualizations. Do customize this report to your needs.

```{r rename-data}
site_species   <- {{site_species}}
species_traits <- {{species_traits}}
site_locations <- {{site_locations}}
```


# Data description

## Species x traits data

The species x traits data consists in `r nrow(species_traits)` species and 
`r ncol(species_traits) - 1` traits (Table \@ref(tab:species-traits) and
Figure \@ref(fig:n-species-by-traits)).


```{r species-traits, results='asis'}
knitr::kable(
  head(species_traits, 10), row.names = FALSE,
  caption = "Extract of species x traits data", longtable = TRUE
)
```

```{r species-traits-description}
fb_table_trait_summary(species_traits, kable = TRUE)
```

The trait correlations are:

```{r trait-correlations, fig.cap='Correlation matrix of traits.'}
fb_plot_trait_correlation(species_traits)
```

## Sites x species data

The sites x species data consists in `r nrow(site_species)` sites and 
`r ncol(species_traits) - 1` species (Table \@ref(tab:site-species)).


```{r site-species, results='asis'}
knitr::kable(
  site_species[1:5, 1:5], row.names = FALSE, 
  caption = "Extract of the site x species data", longtable = TRUE
)
```



## Sites x locations data

The sites x locations data consists in `r nrow(site_locations)` sites.


```{r site-locations}
knitr::kable(
 head(as.data.frame(site_locations)),
 caption = "Extract of the site x location data"
)
```


# Data Completeness

## Number of Species per Trait

```{r n-species-by-traits, fig.cap='Number and proportion of species with non-missing trait values. The traits are ordered from the most complete (at the top) to the least complete (at the bottom).'}
fb_plot_number_species_by_trait(species_traits)
```

## Number of Species with Known Multiple Traits

When we consider multiple traits together we obtain the following:

```{r n-trait-by-species, fig.cap='Number and proportion of species with at least N traits. The graph consider nested categories of species having 0 or more traits.'}
fb_plot_number_traits_by_species(species_traits)
```

## Completeness of Species x Traits matrix

```{r species-traits-completeness, fig.cap='Visualization of the completeness of species x trait matrix. Here the rows represent species and the column the different traits. The columns are ordered from the most complete trait to the least complete one (considering all the traits taken together). Rows are grouped by number of missing traits and shared pattern of missing traits.'}
fb_plot_species_traits_completeness(species_traits)
```

Instead of having rows proportional to the number of species we can also look at combinations of missing and present traits explicitly:

```{r species-traits-completeness-explicit, fig.cap='Observed number and frequencies of combination of missing and present trait values in the species x traits matrix.'}
fb_plot_trait_combination_frequencies(species_traits)
```


## Distribution of Site Trait Coverage

```{r distrib-site-trait-coverage, fig.cap = 'Distribution of trait coverage, i.e. the proportion of species with known trait values, per site for each trait separately.'}
fb_plot_distribution_site_trait_coverage(site_species, species_traits)
```

## Map of Trait Coverage across Sites

```{r map-site-trait-coverage, fig.cap = 'Map of trait coverage, i.e. the proportion of species with known trait values, across sites for each trati separately and all traits taken together'}
fb_map_site_traits_completeness(site_locations, site_species, species_traits)
```


# Session Info

```{r session-info}
sessionInfo()
```