---
title: "{{title}}"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  bookdown::html_document2: 
    toc: true
    toc_float: true
    number_sections: true
    fig_width: 7
    fig_height: 7
    fig_caption: true
    keep_md: false
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: false
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 7
    fig_caption: true
    keep_md: false
  bookdown::word_document2:
    toc: true
    number_sections: true
    fig_width: 7
    fig_height: 7
    fig_caption: true
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width  = 7,
  fig.height = 7,
  dpi        = 100,
  out.width  = "100%",
  echo       = FALSE,
  eval       = TRUE
)
```


\



```{r rename-data}
site_species   <- {{site_species}}
species_traits <- {{species_traits}}
site_locations <- {{site_locations}}
```



# Introduction {-}

Cupidatat nostrud occaecat labore exercitation eu nulla eiusmod veniam eu esse 
nulla commodo fugiat. Proident tempor veniam aliquip exercitation Lorem 
adipisicing labore non officia veniam ipsum amet. Cupidatat et dolor cupidatat 
amet elit labore commodo. Duis reprehenderit duis ullamco fugiat deserunt. 

Irure esse non id ea anim dolor non ullamco sit. Aliquip aliquip nisi consequat 
eu labore cillum incididunt aliquip magna dolor deserunt quis labore.



# Data description

## Species x traits data

The species x traits data consists in `r nrow(species_traits)` species and 
`r ncol(species_traits) - 1` traits (Table \@ref(tab:species-traits) and
Figure \@ref(fig:n-species-by-traits)).


\


```{r species-traits, results='asis'}
knitr::kable(head(species_traits, 10), row.names = FALSE, 
             caption = "Species x traits data", longtable = TRUE)
```



```{r n-species-by-traits, fig.cap='Number of species by trait'}
fb_plot_number_species_by_trait(species_traits)
```



## Sites x species data

The sites x species data consists in `r nrow(site_species)` sites and 
`r ncol(species_traits) - 1` species (Table \@ref(tab:site-species)).


\


```{r site-species, results='asis'}
knitr::kable(head(site_species[ , 1:6], 10), row.names = FALSE, 
             caption = "Sites x species data", longtable = TRUE)
```



## Sites x locations data

The sites x locations data consists in `r nrow(site_locations)` sites.


\


```{r site-locations}
print(site_locations)
```
